---
output: 
  github_document:
    toc: false
    fig_width: 7
    fig_height: 5
---
  
  <!-- README.md is generated from README.Rmd. Please edit that file -->
  
<!-- badges: start -->
[![R-CMD-check](https://github.com/ptompalski/CanadaForestAllometry/actions/workflows/R-CMD-check.yaml/badge.svg?branch=main)](https://github.com/ptompalski/CanadaForestAllometry/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/ptompalski/CanadaForestAllometry/graph/badge.svg)](https://app.codecov.io/gh/ptompalski/CanadaForestAllometry)
<!-- badges: end -->

```{r, echo = FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(CanadaForestAllometry)

options(digits = 2)

knitr::opts_chunk$set(
  collapse = TRUE,
  fig.path = "man/figures/",
  fig.width = 7,
  fig.height = 5,
  comment = "#>",
  dev = "png",
  dpi = 300
)
```
  
# CanadaForestAllometry

`CanadaForestAllometry` is an R package that brings together a collection of published allometric models developed for Canadian forests. The package provides a unified interface to estimate a range of tree- and stand-level attributes, based on models published in the Canadian forestry literature. The focus of `CanadaForestAllometry` is on standardization: models are implemented as faithfully as possible to their original formulations, with transparent parameter tables, consistent inputs/outputs, and jurisdiction-appropriate assumptions (e.g. merchantability rules).

`CanadaForestAllometry` currently includes allometric models to:

- Estimate tree-level volume (total and merchantable)
- Estimate aboveground biomass (AGB)
- Convert volume to biomass
- Convert total volume to merchantable volume
- Apply simple growth models to estimate changes in attributes over time



## Included models

### Aboveground biomass
- `agb_lambert_ung()` -  Canadian national tree aboveground biomass equations (Lambert et al. 2005, Ung et al. 2008).

Example:

```{r, echo=T, warning=F, message=F}
 trees <- tibble::tibble(
  tree_id = 1:4,
  DBH = c(22, 30, 18, 35),
  height = c(18, 24, NA, 27),
  species = c("PICE.GLA", "PICE.GLA", "ABIE.BAL", "PINU.BAN")
 )

trees |>
  dplyr::mutate(
    agb = agb_lambert_ung(
      DBH = DBH,
      height = height,
      species = species,
      keep_model_id = TRUE
    )
  ) |>
  unnest(agb)
```


### National volume-to-biomass conversions models

Implementation of the national volume-to-biomass conversion models (Boudewyn et al. 2007), using the updated model parameters available at https://nfi.nfis.org/en/biomass_models. 

- `v2b()` - converts merchantable volume to biomass components
- `vol_total_to_merchantable()` - converts total volume to merchantable volume
- `agb_component_proportions()` - calculates component proportions (stemwood, bark, branches, foliage) using either merchantable volume or total aboveground biomass as input

### Other
- National growth and yield model (H, BA, V) (Ung et al. 2009)


## Installation 

You can install the most recent version of the package by executing the code below:

```{r, eval=FALSE}
devtools::install_github("ptompalski/CanadaForestAllometry")
library(CanadaForestAllometry)
```

## License

This package is licensed under the GNU Lesser General Public License (LGPL-3.0).

© His Majesty the King in Right of Canada, as represented by the Minister of Natural Resources, 2026.

See the `LICENSE` file for full license terms.

## References 

Boudewyn, P.A.; Song, X.; Magnussen, S.; Gillis, M.D. (2007). Model-based, volume-to-biomass conversion for forested and vegetated land in Canada. Natural Resources Canada, Canadian Forest Service, Pacific Forestry Centre, Victoria, BC. Information Report BC-X-411. 112 p.

Fortin, M., DeBlois, J., Bernier, S., Blais, G., 2007. Mise au point d’un tarif de cubage général pour les forêts québécoises : une approche pour mieux évaluer l’incertitude associée aux prévisions. The Forestry Chronicle 83, 754–765. https://doi.org/10.5558/tfc83754-5


Gal, J., & Bella, I.E. (1994). New stem taper functions for 12 Saskatchewan timber species. Natural Resources Canada, Canadian Forest Service, Northwest Region, Information Report NOR-X-338. Table 5.

Honer, T.G. (1967). Standard volume tables and merchantable conversion actors. Canadian Forest Service Information Report N-X-67.

Honer, T.G.; Ker, M.F.; Alemdag, I.S. 1983. Metric timber tables for the commercial tree species of central and eastern Canada. Environ. Can., Can. For. Serv., Maritimes For. Res. Cent., Fredericton, NB. Inf. Rep. M-X-140. https://ostrnrcan-dostrncan.canada.ca/handle/1845/239814

Huang, S. (1994). Ecologically Based Individual Tree Volume Estimation for Major Alberta Tree Species. Report 1 - Individual tree volume estimation procedures for Alberta: Methods of Formulation and Statistical Foundations. Alberta Environmental Protection, Land and Forest Service, Forest Management Division, Edmonton, AB.

Ker, M.F. 1974. Metric Tree Volume Tables for Newfoundland. Newfoundland Forest Research Centre, St. Johns, NF. Information Report N-X-122. https://ostrnrcan-dostrncan.canada.ca/handle/1845/238893

Kozak, A. (1988). A variable-exponent taper equation. Canadian Journal of Forest Research, 18, 1363–1368

Kozak, A. (1994). Development of Taper Equations by BEC Zones and Species. Province of British Columbia, Ministry of Forests (report).

Klos, R. J., Wang, G. G., Dang, Q.-L., & East, E. W. (2007). Taper equations for five major commercial tree species in Manitoba, Canada. Western Journal of Applied Forestry, 22(3), 163–170.

Lambert, M. C., Ung, C. H., & Raulier, F. (2005). Canadian national tree aboveground biomass equations. Canadian Journal of Forest Research, 35(8), 1996–2018. https://doi.org/10.1139/x05-112

Nigh, G.D., 2016. Total and merchantable volume equations for common tree species in British Columbia: by region and biogeoclimactic zone (No. Prov. B.C., Victoria, B.C. Tech. Rep. 106.).

Sharma, M. (2021). Total and Merchantable Volume Equations for 25 Commercial Tree Species Grown in Canada and the Northeastern United States. Forests, 12, 1270.

Ung, C.H., Bernier, P., & Guo, X.-J. (2008). Canadian national biomass equations: new parameter estimates that include British Columbia data. Canadian Journal of Forest Research, 38(5), 1123–1132. https://doi.org/10.1139/X07-224
 
Ung, C.H., Bernier, P.Y., Guo, X.J., Lambert, M.-C., 2009. A simple growth and yield model for assessing changes in standing volume across Canada’s forests. The Forestry Chronicle 85, 57–64. https://doi.org/10.5558/tfc85057-1

Ung, C.H., Guo, X.J., Fortin, M., 2013. Canadian national taper models. Forestry Chronicle 89, 211–224. https://doi.org/10.5558/tfc2013-040

Warren, G.R. & Meades, J.P. (1986). Wood defect and density studies II: Total and net volume equations for Newfoundland’s forest management units. Canadian Forest Service Information Report N-X-242.

Zakrzewski, W.T., Penner, M., 2013. A comparison of tree stem taper models for use in Ontario. Ontario Forest Research Institute, Report 176.
