---
title: "Tree Aboveground Biomass Models"
bibliography: ../../REFERENCES.bib
csl: ../../isprs-journal-of-photogrammetry-and-remote-sensing.csl
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
#| echo: false
#| warning: false
#| message: false
library(CanadaForestAllometry)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
```

The `CanadaForestAllometry` package implements the national aboveground biomass (AGB) equations originally developed by @Lambert2005 and subsequently revised by @Ung2008. The revision by @Ung2008 incorporated previously unavailable British Columbia data, expanding species coverage and updating parameter estimates while preserving the original additive modelling framework.

These equations were derived from destructively sampled trees across Canada and estimate dry biomass for four aboveground components: stem wood, stem bark, branches, and foliage. Two model formulations were developed: one using DBH alone, and a second using DBH combined with total tree height.

The `agb_lambert_ung()` function implements both equation systems within a single interface. When a valid height is supplied, the DBH–height formulation is used; otherwise, the DBH-only formulation is applied.



## Example:

```{r, echo=T, warning=F, message=F}
 trees <- tibble::tibble(
  tree_id = 1:4,
  DBH = c(22, 30, 18, 35),
  height = c(18, 24, NA, 27),
  species = c("PICE.GLA", "PICE.GLA", "ABIE.BAL", "PINU.BAN")
 )

trees |>
  dplyr::mutate(
    agb = agb_lambert_ung(
      DBH = DBH,
      height = height,
      species = species,
      keep_model_id = TRUE
    )
  ) |>
  unnest(agb)
```


# References